---
title: "STAT0218 Final Project"
author: "Ellie Suit, Andy Atallah, and Ai Hattori"
date: "2023-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = F, message = F}
# libraries 
library(tidyverse)
library(survival)
library(rpart)
library(rpart.plot)
library(dendextend) # hierarchical clustering
library(lubridate) # hierarchical clustering
library(kernlab) # SVM
library(e1071) # SVM
```


```{r, warning = F, message = F}
library(randomForest)
library(caret)
trees <- read_csv("Tree_Data.csv")
```

# EDA: 

Researchers conducted a field experiment, "consisting of four tree species, seven soil sources (sterilized conspecific, live conspecific, and five heterospecific), and a gradient of forest understory light levels (low, medium, and high)." 

```{r}
ggplot(data = trees, mapping = aes(x = Species)) + 
  geom_bar(fill = "lightblue") +
  labs(y = "Count", x = "Tree species",
       title = "Number of trees by species")
```

```{r}
tree_EDA <- trees %>%
  mutate(SoilSource = paste(Sterile, ",", Conspecific))

ggplot(data = tree_EDA, mapping = aes(x = SoilSource, fill = Species)) + 
  # geom_bar(position = "fill") +
  geom_bar() +
  labs(y = "Count", x = "Soil source",
       title = "Soil sources by species")
```

How to manually order x-axis on bar chart: https://community.rstudio.com/t/how-to-manually-order-x-axis-on-bar-chart/9601 was used for the following code.
```{r}
trees %>% 
  mutate(Light_Cat = factor(Light_Cat, 
                                    levels = c("Low", "Med", "High"))) %>% 
  ggplot(mapping = aes(x = Light_Cat, fill = Species)) +
  geom_bar() + 
  labs(y = "Count", x = "Light level",
       title = "Light levels by species")
```
#  Unsupervised learning: Hierarchical clustering

How to select only numeric columns: https://www.geeksforgeeks.org/select-only-numeric-columns-from-dataframe-in-r/ 

How to convert non-standard format of dates to standard format (YYYY-MM-DD): https://epirhandbook.com/en/working-with-dates.html

Below, we conducted hierarchical clustering for numerical variables. Among the numerical variables, we did not include the following variables for the following reasons. 

* No: because this is an ID number unique to each tree

* Census: This is a "census number at which time the seedling died or was harvested." Since we do not have any information about what the census number means, we excluded the variable for all the analyses we did for this project. Hence, we did not include this variable in hierarchical clustering. 

```{r, warning = F}
# convert PlantDate in character format to date time object
tree_hc <- trees %>% 
  mutate(PlantDate = mdy(PlantDate)) %>% # change to standard format(Year-Month-Day)
  mutate(PlantDateYear = year(PlantDate),
          PlantDateMonth = month(PlantDate),
          PlantDateDay = day(PlantDate))

# scale tree data without the variable Census
tree_scaled1 <- select_if(tree_hc, is.numeric) %>% # select only numeric columns
                select(-c(No, Census)) 
               
# calculate all pairwise distances
tree_distances1 <- dist(tree_scaled1)

# Do hierarchical clustering
hc1 <- tree_distances1 %>%
  hclust() %>%
  as.dendrogram() %>%
  place_labels(tree$No) 

plot(hc1) 
title('Dendrogram without the variable Census')

# scale tree data with the variable Census
tree_scaled2 <- select_if(tree_hc, is.numeric) %>% # select only numeric columns
                select(-c(No))
     
# calculate all pairwise distances
tree_distances2 <- dist(tree_scaled2)

# Do hierarchical clustering
hc2 <- tree_distances2 %>%
  hclust() %>%
  as.dendrogram() %>%
  place_labels(tree$No) 

plot(hc2) 
#title('Dendrogram with the variable Census')

# Cut this dendrogram at a certain height to identify outliers
# cuts <- cutree(hc, h=120)
# # 
# tree_with_clusters <- trees %>%
#    mutate(group = cuts)
# # 
# tree_with_clusters %>%
#    filter(group == 2)
```

There is no split with only one terminal node in both dendrograms with or without the variable Census. 

# PCA (and SVM for prediction)

Use PCA to build a model that predicts the variable Time (the number of days at which time the seedling died or was harvested) by using the following variables:

* Light ISF: the amount of light reaching each subplot at a height of 1m
* Core: Year the soil core was removed from the field
* AMF: Percent arbuscular mycorrhizal fungi colonization on the fine roots of harvested seedlings
* EMF: Percent ectomycorrhizal fungi colonization on the root tips of harvested seedlings
* Phenolics: nmol Gallic acid equivalents per mg dry extract
* NSC: percent dry mass nonstructural carbohydrates
* Lignin: percent dry mass lignin

Because PCA only works with numerical variables, I had to remove the following variables but I think they should ideally be used for prediction:

** Species
** Soil: Species from which the soil core was taken 
** Adult: individual tree that soil was taken from
** Sterile: whether the soil was sterilized or not
** Conspecific: whether the soil was conspecific, heterospecific, or sterlized conspecific
** Myco: Mycorrhizal type of the seedling species (AMF or EMF)
** SoilMyco: Mycorrhizal type of the species culturing the soil (AMF or EMF)


(Variables which I think were fine to be removed: No, Plot, Subplot, Light Cat, PlantDate, Census, Event, Harvest, Alive - it did not make sense to include all of these given that our goal was to build a model that predicts Time)

## Find loading score 
```{r}
# select necessary variables
tree_pca <- trees %>%
  select(-c(No, Plot, Subplot, Light_Cat, PlantDate, Census, Event, Harvest, Alive)) %>%
  select_if(is.numeric) %>%
  na.omit()

# scale data and conduct PCA
pca1 <- prcomp(tree_pca %>% select(-Time), scale = TRUE) # remove what you are predicting (Time)

# the first PC (direction of the data along which the observations vary the most)
sort(pca1$rotation[,1])
```
## Build a model to predict Time by using loading score

SVM with Caret: https://rpubs.com/uky994/593668

```{r}
set.seed(1)

# create a data frame with Time and loading score from the first PC
tree_pca_subset <- data.frame(loading_score = pca1$x[,1]) %>% # score in PC1 for each tree
  mutate(Time = tree_pca$Time)

# model with linear kernel
pca_model <- train(Time ~ loading_score,
                  data = tree_pca_subset,
                  method = "svmLinear",
                  tuneGrid = expand.grid(C = c(0.25, 0.5, 1)),
                  coob = TRUE) # calculate out-of-bag error
pca_model

# find the smallest RMSE
pca_model$results[which.min(pca_model$results[,2]),]
```

With linear kernel, the smallest RMSE is approximately 41.2.

```{r}
set.seed(1)
# model with radial kernel
pca_model2 <- train(Time ~ loading_score,
                  data = tree_pca_subset,
                  method = "svmRadial",
                  coob = TRUE) # calculate out-of-bag error
pca_model2

# find the smallest RMSE
pca_model2$results[which.min(pca_model2$results[,3]),]
```

With radial kernel, the smallest RMSE is approximately 43.5.

```{r}
set.seed(1)
# model with polynomial kernel
pca_model3 <- train(Time ~ loading_score,
                  data = tree_pca_subset,
                  method = "svmPoly",
                  coob = TRUE) # calculate out-of-bag error
pca_model3

# find the smallest RMSE
pca_model3$results[which.min(pca_model3$results[,4]),]
```
With polynomial basis kernel, the smallest RMSE is 40.6.

# Supervised learning: SVM

Use multiple variables (instead of the single variable, PCA loading score) for SVM. 

```{r}
set.seed(1)
tree_svm <- trees %>%
  select(-c(No, Plot, Subplot, Light_Cat, PlantDate, Census, Event, Harvest, Alive)) %>%
  mutate(Species = factor(Species),
         Soil = factor(Soil),
         Adult = factor(Adult),
         Sterile = factor(Sterile),
         Conspecific = factor(Conspecific),
         # Myco = factor(Myco),
         SoilMyco = factor(SoilMyco)) %>%
  select(-c(AMF, EMF)) %>% # including AMF and EMF results in losing all observations with # Myco = AMF
  na.omit()

svm_model = train(Time ~ .,
                  data = tree_svm,
                  method = "svmLinear",
                  tuneGrid = expand.grid(C = c(0.25, 0.5, 1)),
                  coob = TRUE)
svm_model

# find the smallest RMSE
svm_model$results[which.min(svm_model$results[,2]),]
```
SVM with variables in the dataset (instead of the loading score) results in the smallest RMSE of around 30.6.


# Ellie's part
Remove unwanted variables
```{r}
library(lubridate)
trees_clean <- trees %>%
  select(-c(Adult, EMF, AMF, Alive, Harvest)) %>%
  mutate(Subplot = factor(Subplot)) %>%
  mutate(Species = factor(Species)) %>%
  mutate(Soil = factor(Soil)) %>%
  mutate(Sterile = factor(Sterile)) %>%
  mutate(Conspecific = factor(Conspecific)) %>%
  mutate(Myco = factor(Myco)) %>%
  mutate(SoilMyco = factor(SoilMyco)) %>%
  mutate(PlantDate = as.Date(PlantDate, format="%m/%d/%Y")) %>%
  na.omit() 
#View(trees_clean)
  
#trees_clean %>%
 #filter(if_any(everything(), is.na))



rows <- seq(from = 1, to = nrow(trees), by = 1)
row_nums <- as.data.frame(rows)

trees_clean2 <- cbind(row_nums, trees_clean)
```



Create a single survival tree

```{r}
surv_tree <- rpart(Surv(Time, Event) ~., data = trees_clean)

rpart.plot(surv_tree)
```



Results
- Most important variables: Census, Phenolics, certain categories of adult trees, AMF
```{r}
# Calculate confusion matrix to summarize these results


```


```{r}
library(ranger)
rf <- ranger(Surv(Time, Event) ~.-No, data = trees_clean, importance = 'impurity')


as.data.frame(rf$variable.importance) %>%
  arrange(-rf$variable.importance)


```










